STJ2012 analysis

### 2016-08-14 analysis/STJ2012_analysis.Rmd
move R functions to separate scripts

#### Descriptive statistics and distribution of OTU counts
```{r phy_stats, cache=T, fig.height=4, fig.width=8}
# Compute summary statistics
stats <- function(phy) {
  return(
    data.frame(
      'Total count in OTU table'=sum(otu_table(phy)),
      'Number of OTUs'=ntaxa(phy),
      'Range of OTU counts'=paste0(range(taxa_sums(phy)), collapse=" - "),
      'Number of singleton OTUs'=length(taxa_sums(phy)[taxa_sums(phy) <= 1]),
      'Number of samples'=nsamples(phy),
      'Range of reads per sample'=paste0(range(sample_sums(phy)), collapse=" - "),
      'Arithmetic mean (±SD) reads per sample'=paste(as.integer(mean(sample_sums(phy))), 
                                                     as.integer(sd(sample_sums(phy))), sep=" ± "),
      'Geometric mean (±SD) reads per sample'=paste(as.integer(exp(mean(log(sample_sums(phy))))),
                                                    as.integer(exp(sd(log(sample_sums(phy))))), sep=" ± "),
      check.names=F)
  )
}
stats97 <- data.frame(`97% OTUs`=t(stats(phy97)), check.names=F)
stats97bs <- data.frame(`97% within-sample OTUs`=t(stats(phy97bs)), check.names=F)
stats100 <- data.frame(`100% OTUs`=t(stats(phy100)), check.names=F)

@@

#### Filtered data: descriptive statistics and distribution of OTUs
```{r phy.f_histograms, fig.height=4, fig.width=8}
# Compute summary statistics
stats97.f <- data.frame(`97% OTUs`=t(stats(phy97.f)), check.names=F)
stats97bs.f <- data.frame(`97% within-sample OTUs`=t(stats(phy97bs.f)), check.names=F)
stats100.f <- data.frame(`100% OTUs`=t(stats(phy100.f)), check.names=F)

@@

# Data analysis
## Community composition barplots
Here the composition of each sample is plotted as a horizontal bar, sorted by species and shore. Each segment of the bar represents an individual OTU. In the lefthand plot, all OTUs are colored by clade, in the right OTU, all OTUs are given a unique color.
```{r cladecomp97, cache=T, fig.height=6, fig.width=8}
# Define function to plot Symbiodinium community composition
composition <- function(phy, col, legend=T) {
  samdat <- data.frame(sample_data(phy))
  samdat$Genus <- factor(samdat$Genus, levels=rev(levels(samdat$Genus)))
  samdat$Species <- factor(samdat$Species, levels=rev(levels(samdat$Species)))
  samdat$Shore <- factor(samdat$Shore, levels=rev(levels(samdat$Shore)))
  samdat <- samdat[with(samdat, order(Genus, Species, Shore)), ]
  typerelabund <- as.matrix(otu_table(phy)[order(data.frame(tax_table(phy))$Subtype), 
                                           rownames(samdat)])
  shorebreaks <- c(as.character(samdat$Shore), "X")==c("X", as.character(samdat$Shore))
  shorebreaks <- which(shorebreaks==F) - 1
  spbreaks <- c(which(duplicated(samdat$Species)==F) - 1, nrow(samdat))
  # Make Barplot
  barplot(typerelabund, horiz=T, space=0, axes=F,axisnames=F, yaxs="i", col=col)
  rect(0, 0, par("usr")[2], par("usr")[4], lwd=1, xpd=T)
  axis(side=1, at=seq(0, 1, 0.1), line=0, tck=-0.025, mgp=c(0,0.25,0), cex.axis=0.7)
  mtext(side=1, "Relative abundance", cex=0.7, line=1)
  # Add legend
  if (legend==T) {
    legend(x=par("usr")[2]/2, y=par("usr")[4], xjust=0.5, yjust=0.25, horiz=T, bty="n", xpd=T, 
           cex=0.7, legend=c("A", "B", "C", "D", "F", "G"), fill=taxcolors, x.intersp=0.5)
    legend(x=par("usr")[2]-0.03, y=par("usr")[4], xjust=0, yjust=0.1, bty="n", xpd=T, cex=0.6, 
           pt.cex=0.6, legend=c("N Shore", "S Shore"), fill=c(0, "gray40"), y.intersp=0.7, 
           x.intersp=0.3)
  }
  # Add grouping bars for Shore
  for (i in 1:length(shorebreaks)) {
    lines(c(0, 1), c(shorebreaks[i], shorebreaks[i]), lty=2, lwd=0.25)
    rect(1.01, shorebreaks[i], 1.04, shorebreaks[i+1], col=rep(c("gray40", 0), 50)[i], 
         lwd=0.25, xpd=T)
  }
  # Add lines to separate species and species names
  for (i in 1:length(spbreaks)) {
    lines(c(0, 1.07), c(spbreaks[i], spbreaks[i]), xpd=T, type="l", lwd=0.4)
    text(1.03, (spbreaks[i] + spbreaks[i+1]) / 2, xpd=T, pos=4, cex=0.6,
        labels=paste(samdat$Genus[which(duplicated(samdat$Species)==F)][i], "\n",
                    samdat$Species[which(duplicated(samdat$Species)==F)][i], sep=""))
  }
}

@@

```{r shorediffs, cache=T}
shorediffs <- function(phy) {
  # Compute within- and between-shore statistics for each Species
  shorestats <- data.frame(matrix(dimnames=list(levels(data.frame(sample_data(phy))$Species), 
                                                c("Species", "n", "overall", "within", 
                                                  "between", "R2", "bd", "p")), ncol=8, nrow=10))
  for (i in 1:nlevels(data.frame(sample_data(phy))$Species)) {
    sp <<- levels(data.frame(sample_data(phy))$Species)[i]
    shorestats$Species[i] <- sp
    phy.species <- subset_samples(phy, Species==sp)  # Set temp phyloseq object
    shorestats$n[i] <- nsamples(phy.species)
    md <- meandist(phyloseq::distance(phy.species, "bray"), 
                   sample_data(phy.species)$Shore)
    shorestats$within[i] <- summary(md)$W  # Weighted mean dissimilarity within both Pools
    shorestats$between[i] <- summary(md)$B  # Mean dissimilarity between Pools
    shorestats$overall[i] <- summary(md)$D  # Overall dissimilarity
    if (nlevels(as(sample_data(phy.species), "data.frame")$Shore) > 1) {
      #set.seed(70235)
      set.seed(43789)
      permanova <- adonis(phyloseq::distance(phy.species, "bray") ~ Shore, 
                          data=as(sample_data(phy.species), "data.frame"), permutations=999)
      shorestats$R2[i] <- permanova$aov.tab$"R2"[1]  # PERMANOVA partial R-squared
      shorestats$p[i] <- permanova$aov.tab$"Pr(>F)"[1]  # PERMANOVA p-value
      bd <- betadisper(phyloseq::distance(phy.species, "bray"),
                       group=as(sample_data(phy.species), "data.frame")$Shore)
      shorestats$bd[i] <- TukeyHSD(bd)$group[4]
    } else {
      shorestats$R2[i] <- NA
      shorestats$p[i] <- NA
      shorestats$bd[i] <- NA
    }
  }
  return(shorestats)
}

shorestats97 <- shorediffs(phy97.f.t)
shorestats97bs <- shorediffs(phy97bs.f.t)
shorediffs(phy97bs.f.p)
shorestats100 <- shorediffs(phy100.f.t)

@@

## Beta diversity
Beta diversity is evaluated here as the multivariate dispersion of samples within a coral species group. Principal coordinate analysis on Bray-Curtis dissimilarities is used to calculate average distance-to-centroid values for each species group, which are then compared statistically by a permutation test. This analysis is implemented using betadisper() in the vegan package, based on [Anderson (2006)](dx.doi.org/10.1111/j.1541-0420.2005.00440.x).
```{r betadisper, cache=T, fig.width=10, fig.height=4}
betad <- function(phy) {
  # Calculate distances to species centroids for each sample in principal coordinate space
  sambd <- betadisper(d=vegdist(t(otu_table(phy)), method="bray"), 
                      group=data.frame(sample_data(phy))$Species, type="centroid",
                      bias.adjust=TRUE)
  # Calculate each species' mean distance to centroid and standard error
  sambdsumm <- aggregate(data.frame(mean=sambd$distances), by=list(group=sambd$group), FUN=mean)
  sambdsumm$se <- aggregate(sambd$distances, by=list(group=sambd$group), 
                            FUN=function(x) sd(x)/sqrt(length(x)))$x
  # Determine order of decreasing mean dispersion
  sambdsumm.ord <- sambdsumm[order(sambdsumm$mean, decreasing=T), ]
  # Update betadisper object with species in decreasing order of mean dispersion
  sambd <- betadisper(d=vegdist(t(otu_table(phy)), method="bray"), 
                      group=factor(data.frame(sample_data(phy))$Species,
                                   levels=as.character(sambdsumm.ord$group)), 
                      type="centroid", bias.adjust=TRUE)
  # Use permutations to perform pairwise comparisons of group mean dispersions
  sampt <- permutest(sambd, pairwise=T)
  saml <- multcompLetters(sampt$pairwise$permuted)
  # Return results as a list
  return(list(sambdsumm.ord=sambdsumm.ord, sambdsumm=sambdsumm, saml=saml))
}

betad97 <- betad(phy97.f.t)
betad97bs <- betad(phy97bs.f.t)
betad100 <- betad(phy100.f.t)

@@

```{r otu_barplots_custom, fig.width=8, fig.height=8}
otubarplot <- function(phy, main) {
  samdat <- data.frame(sample_data(phy))
  typerelabund <- as.matrix(otu_table(phy)[order(data.frame(tax_table(phy))$Subtype), 
                                                 rownames(samdat)])
  typerelabund <- typerelabund[, rownames(sample_data(phy)[with(sample_data(phy), 
                                                                order(Shore, Site, InputFileName))])]
  # Get info for plotting OTU names and blast hits on top of barplot
  blasthits <- as.character(data.frame(tax_table(phy))[order(data.frame(tax_table(phy))$Subtype), "Subtype"])
  names <- as.character(rownames(data.frame(tax_table(phy)))[order(data.frame(tax_table(phy))$Subtype)])
  samples <- colnames(typerelabund)
  shores <- data.frame(sample_data(phy))[samples, "Shore"]
  sites <- data.frame(sample_data(phy))[samples, "Site"]
  divides <- apply(typerelabund, 2, cumsum)
  heights <- diff(divides)
  heights[heights < 0.04] <- NA
  # Plot barplot and OTU names and blast hits
  bars <- barplot(typerelabund, col=rainbow(ntaxa(phy)), las=1, space=0.1, xaxs="i",
                  xlab="", ylab="", xaxt="n")
  text(bars, -0.06, labels=samples, xpd=T)
  text(bars, -0.10, labels=shores, xpd=T)
  text(bars, -0.14, labels=substr(sites, 1, 2), xpd=T)
  text(bars[1]/4, c(-0.06, -0.10, -0.14), labels=c("sample:", "shore:", "site:"), xpd=T, pos=2)
  mtext(side=2, text = "Relative Abundance", line=2)
  for (i in 1:length(bars)) {
    text(rep(bars[i], length(heights[which(!is.na(heights[,i])),i])), 
         divides[which(!is.na(heights[,i])),i] + heights[which(!is.na(heights[,i])),i] / 2, 
         labels=paste(names[which(!is.na(heights[,i]))+1],blasthits[which(!is.na(heights[,i]))+1],sep="\n"),
         cex=0.75)
  }
  mtext(side=3, main, xpd=T, adj=0, line=0.5, font=2)
}
```

@@

# Network analysis from tutorial
```{r networks}
make.net <- function(physeq, n) {
  # Remove taxa that are not present in subset
  physeq <- prune_taxa(rowSums(otu_table(physeq))!=0, physeq)
  # Convert OTU table to "edges" table (weight = relative abundance)
  otudf <- data.frame(t(otu_table(physeq)), check.names=F)
  edges <- setNames(melt(as.matrix(otudf)), nm=c("source", "target", "weight"))  # Melt data
  edges <- edges[edges$weight>n, ]
  # Create "nodes" table from sample and tax data
  sdf <- data.frame(sample_data(physeq))
  sdf$id <- rownames(sdf)
  tdf <- data.frame(tax_table(physeq))
  tdf$id <- rownames(tdf)
  nodes <- merge(sdf, tdf, all=T)
  # Create network
  net <- graph_from_data_frame(d=edges, vertices=nodes, directed=F)
  return(net)
}
dstr.net <- make.net(dstr97bs.f.p, 0)

@@

pfur.net <- make.net(pfur97bs.f.p, 0)
plot(pfur.net,
     edge.curved=0.1, 
     edge.width=10*(E(pfur.net)$weight)^0.2,
     vertex.label=substr(V(pfur.net)$Subtype, 1, 4),
     vertex.color=taxcolors[factor(V(pfur.net)$Clade, levels=c("A","B","C","D","F","G"))])

oann.net <- make.net(oann97bs.f.p, 0)
plot(oann.net,
     edge.curved=0.1, 
     edge.width=10*(E(oann.net)$weight)^0.2,
     vertex.label=substr(V(oann.net)$Subtype, 1, 4),
     vertex.color=taxcolors[factor(V(oann.net)$Clade, levels=c("A","B","C","D","F","G"))])

malc.net <- make.net(malc97bs.f.p, 0)
plot(malc.net,
     edge.curved=0.1, 
     edge.width=10*(E(malc.net)$weight)^0.2,
     vertex.label=substr(V(malc.net)$Subtype, 1, 4),
     vertex.color=taxcolors[factor(V(malc.net)$Clade, levels=c("A","B","C","D","F","G"))])

set.seed(9147)
ssid.net <- make.net(ssid97bs.f.p, 0)
plot(ssid.net,
     edge.curved=0.1, 
     edge.width=10*(E(ssid.net)$weight)^0.2,
     vertex.label=substr(V(ssid.net)$Subtype, 1, 4),
     vertex.color=taxcolors[factor(V(ssid.net)$Clade, levels=c("A","B","C","D","F","G"))])

ffra.net <- make.net(ffra97bs.f.p, 0)
plot(ffra.net,
     edge.curved=0.1, 
     edge.width=10*(E(ffra.net)$weight)^0.2,
     vertex.label=substr(V(ffra.net)$Subtype, 1, 4),
     vertex.color=taxcolors[factor(V(ffra.net)$Clade, levels=c("A","B","C","D","F","G"))])

srad.net <- make.net(srad97bs.f.p, 0)
plot(srad.net,
     edge.curved=0.1, 
     edge.width=10*(E(srad.net)$weight)^0.2,
     vertex.label=substr(V(srad.net)$Subtype, 1, 4),
     vertex.color=taxcolors[factor(V(srad.net)$Clade, levels=c("A","B","C","D","F","G"))])

past.net <- make.net(past97bs.f.p, 0)
plot(past.net,
     edge.curved=0.1, 
     edge.width=10*(E(past.net)$weight)^0.2,
     vertex.label=substr(V(past.net)$Subtype, 1, 4),
     vertex.color=taxcolors[factor(V(past.net)$Clade, levels=c("A","B","C","D","F","G"))])

dcyl.net <- make.net(dcyl97bs.f.p, 0)
plot(dcyl.net,
     edge.curved=0.1, 
     edge.width=10*(E(dcyl.net)$weight)^0.2,
     vertex.label=substr(V(dcyl.net)$Subtype, 1, 4),
     vertex.color=taxcolors[factor(V(dcyl.net)$Clade, levels=c("A","B","C","D","F","G"))])

mcav.net <- make.net(mcav97bs.f.p, 0)
plot(mcav.net,
     edge.curved=0.1, 
     edge.width=10*(E(mcav.net)$weight)^0.2,
     vertex.label=substr(V(mcav.net)$Subtype, 1, 4),
     vertex.color=taxcolors[factor(V(mcav.net)$Clade, levels=c("A","B","C","D","F","G"))])

>>>>

#### Descriptive statistics and distribution of OTU counts
```{r phy_stats, cache=T, fig.height=4, fig.width=8}
# Compute summary statistics
source("R/phystats.R")

stats97 <- data.frame(`97% OTUs`=t(phystats(phy97)), check.names=F)
stats97bs <- data.frame(`97% within-sample OTUs`=t(phystats(phy97bs)), check.names=F)
stats100 <- data.frame(`100% OTUs`=t(phystats(phy100)), check.names=F)

@@

#### Filtered data: descriptive statistics and distribution of OTUs
```{r phy.f_histograms, fig.height=4, fig.width=8}
# Compute summary statistics
stats97.f <- data.frame(`97% OTUs`=t(phystats(phy97.f)), check.names=F)
stats97bs.f <- data.frame(`97% within-sample OTUs`=t(phystats(phy97bs.f)), check.names=F)
stats100.f <- data.frame(`100% OTUs`=t(phystats(phy100.f)), check.names=F)

@@

# Data analysis
## Community composition barplots
Here the composition of each sample is plotted as a horizontal bar, sorted by species and shore. Each segment of the bar represents an individual OTU. In the lefthand plot, all OTUs are colored by clade, in the right OTU, all OTUs are given a unique color.
```{r cladecomp97, cache=T, fig.height=6, fig.width=8}
# Define function to plot Symbiodinium community composition
source("R/composition.R")

@@

```{r shorediffs, cache=T}
source("R/perms.R")
set.seed(43789)
shorestats97 <- perms(phy97.f.t, group="Species", trt="Shore")
shorestats97bs <- perms(phy97bs.f.t, group="Species", trt="Shore")
shorestats100 <- perms(phy100.f.t, group="Species", trt="Shore")

@@

## Beta diversity
Beta diversity is evaluated here as the multivariate dispersion of samples within a coral species group. Principal coordinate analysis on Bray-Curtis dissimilarities is used to calculate average distance-to-centroid values for each species group, which are then compared statistically by a permutation test. This analysis is implemented using betadisper() in the vegan package, based on [Anderson (2006)](dx.doi.org/10.1111/j.1541-0420.2005.00440.x).
```{r betadisper, cache=T, fig.width=10, fig.height=4}
source("R/betad.R")

betad97 <- betad(phy97.f.t, group="Species")
betad97bs <- betad(phy97bs.f.t, group="Species")
betad100 <- betad(phy100.f.t, group="Species")

@@

@@

# Network analysis from tutorial
```{r networks}

dstr.net <- makenet(dstr97bs.f.p, 0)

@@

pfur.net <- makenet(pfur97bs.f.p, 0)
plot(pfur.net,
     edge.curved=0.1, 
     edge.width=10*(E(pfur.net)$weight)^0.2,
     vertex.label=substr(V(pfur.net)$Subtype, 1, 4),
     vertex.color=taxcolors[factor(V(pfur.net)$Clade, levels=c("A","B","C","D","F","G"))])

oann.net <- makenet(oann97bs.f.p, 0)
plot(oann.net,
     edge.curved=0.1, 
     edge.width=10*(E(oann.net)$weight)^0.2,
     vertex.label=substr(V(oann.net)$Subtype, 1, 4),
     vertex.color=taxcolors[factor(V(oann.net)$Clade, levels=c("A","B","C","D","F","G"))])

malc.net <- makenet(malc97bs.f.p, 0)
plot(malc.net,
     edge.curved=0.1, 
     edge.width=10*(E(malc.net)$weight)^0.2,
     vertex.label=substr(V(malc.net)$Subtype, 1, 4),
     vertex.color=taxcolors[factor(V(malc.net)$Clade, levels=c("A","B","C","D","F","G"))])

set.seed(9147)
ssid.net <- makenet(ssid97bs.f.p, 0)
plot(ssid.net,
     edge.curved=0.1, 
     edge.width=10*(E(ssid.net)$weight)^0.2,
     vertex.label=substr(V(ssid.net)$Subtype, 1, 4),
     vertex.color=taxcolors[factor(V(ssid.net)$Clade, levels=c("A","B","C","D","F","G"))])

ffra.net <- makenet(ffra97bs.f.p, 0)
plot(ffra.net,
     edge.curved=0.1, 
     edge.width=10*(E(ffra.net)$weight)^0.2,
     vertex.label=substr(V(ffra.net)$Subtype, 1, 4),
     vertex.color=taxcolors[factor(V(ffra.net)$Clade, levels=c("A","B","C","D","F","G"))])

srad.net <- makenet(srad97bs.f.p, 0)
plot(srad.net,
     edge.curved=0.1, 
     edge.width=10*(E(srad.net)$weight)^0.2,
     vertex.label=substr(V(srad.net)$Subtype, 1, 4),
     vertex.color=taxcolors[factor(V(srad.net)$Clade, levels=c("A","B","C","D","F","G"))])

past.net <- makenet(past97bs.f.p, 0)
plot(past.net,
     edge.curved=0.1, 
     edge.width=10*(E(past.net)$weight)^0.2,
     vertex.label=substr(V(past.net)$Subtype, 1, 4),
     vertex.color=taxcolors[factor(V(past.net)$Clade, levels=c("A","B","C","D","F","G"))])

dcyl.net <- makenet(dcyl97bs.f.p, 0)
plot(dcyl.net,
     edge.curved=0.1, 
     edge.width=10*(E(dcyl.net)$weight)^0.2,
     vertex.label=substr(V(dcyl.net)$Subtype, 1, 4),
     vertex.color=taxcolors[factor(V(dcyl.net)$Clade, levels=c("A","B","C","D","F","G"))])

mcav.net <- makenet(mcav97bs.f.p, 0)
plot(mcav.net,
     edge.curved=0.1, 
     edge.width=10*(E(mcav.net)$weight)^0.2,
     vertex.label=substr(V(mcav.net)$Subtype, 1, 4),
     vertex.color=taxcolors[factor(V(mcav.net)$Clade, levels=c("A","B","C","D","F","G"))])
